---
title: "EDA Lesson 4"
author: "Tom Gocken"
date: "Sunday, January 24, 2016"
output:
  word_document:
    fig_width: 7
---

### Scatterplots
```{r}
# https://s3.amazonaws.com/udacity-hosted-downloads/ud651/EDA_Course_Materials.zip
# Set directory, load data and ggplot2 library

setwd("C:/Projects/UD651")
list.files()
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
library(ggplot2)

qplot(x = age, y = friend_count, data = pf)

# Alternative syntax
qplot(age, friend_count, data = pf)

# ggplot syntax
ggplot(aes(x = age, y = friend_count), data = pf) + geom_point()
summary(pf$age)

# Limit x axis range
ggplot(aes(x = age, y = friend_count), data = pf) + geom_point() + xlim(13,90)
```

### Overplotting
```{r}
# Reduce overplotting with alpha where 1 graph point equal to 20 data points
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20)  + 
  xlim(13,90)

# Add jitter
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_jitter(alpha = 1/20)  + 
  xlim(13,90)

```

### coord_trans
```{r}
# Square root transformation
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20)  + 
  xlim(13,90) +
  coord_trans(y = 'sqrt')

# Jitter age, > 0 only
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0))  + 
  xlim(13,90) +
  coord_trans(y = 'sqrt')
```

### Alpha and Jitter
```{r}
names(pf)

ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  geom_point()

# Add jitter
ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  geom_jitter() +
  xlim(13,90)

# Add alpha
ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  geom_jitter(alpha = 1/10) +
  xlim(13,90)

# Alt jitter syntax, sqrt transformation
ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  geom_point(alpha = 1/10, position = position_jitter(h = 0)) +
  xlim(13,90) +
  coord_trans(y = 'sqrt')
```

### Conditional Means
```{r}
# install.packages('dplyr')
library(dplyr)

age_groups <- group_by(pf, age)

pf.fc_by_age <- summarise(age_groups,
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())

pf.fc_by_age <- arrange(pf.fc_by_age, age)

head(pf.fc_by_age)

# Alternative syntax, note: %.% equivalent to %>%
pf.fc_by_age <- pf %.% 
  group_by(age) %.%
  summarise(friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %.%
  arrange(age)

head(pf.fc_by_age)

# geom_line on summarized values
ggplot(aes(x= age, y = friend_count_mean), data = pf.fc_by_age) +
  geom_line()
```

### Overlaying Summaries with Raw Data
```{r}

# Display raw data and mean
ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  geom_point(alpha = 0.05, 
             position = position_jitter(h = 0),
             color = 'orange') +
  xlim(13,90) +
  coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = mean)

# Add quantiles
ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  geom_point(alpha = 0.05, 
             position = position_jitter(h = 0),
             color = 'orange') +
  xlim(13,90) +
  coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args=list(probs=0.1),
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args=list(probs=0.5), # = median
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args=list(probs=0.9),
            linetype = 2, color = 'blue')


# Use coord_cartesian to zoom in
ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  coord_cartesian(xlim = c(13, 70), ylim = c(0, 1000)) +
  geom_point(alpha = 0.05, 
             position = position_jitter(h = 0),
             color = 'orange') +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args=list(probs=0.1),
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args=list(probs=0.5), # = median
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args=list(probs=0.9),
            linetype = 2, color = 'blue')
```

### Correlation
```{r}
cor.test(pf$age, pf$friend_count, method = 'pearson')

# Alt syntax
with(pf, cor.test(age, friend_count, method = 'pearson'))
```

### Correlation on Subsets
```{r}
with(subset(pf, age <= 70), cor.test(age, friend_count)) # method = 'peason' is default
```

### Create Scatterplots
```{r}
ggplot(aes(x = www_likes_received, likes_received), data = pf) +
  geom_point()
```

### Strong Correlations
```{r}
# Limit range using 95% quantiles, add line for correlation coefficient
ggplot(aes(x = www_likes_received, likes_received), data = pf) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_smooth(method = 'lm', color = 'red')

cor.test(pf$www_likes_received, pf$likes_received)
# High correlation not meaningful since www_likes_received is a subset of likes_received
```

### More Caution with Correlation
```{r}
# install.packages('alr3')
library(alr3)
data(Mitchell)
# ?Mitchell
# Mitchell soil temperature
# Description: Data collected by Kenneth G. Hubbard on soil temperature at 20 cm depth 
# in Mitchell, Nebraska for 17 years (1976-1992) The variable month is the month number.

ggplot(data = Mitchell, aes(x = Month, y = Temp)) +
  geom_point()

# Alt syntax
qplot(data = Mitchell, Month, Temp)
```

### Noisy Scatterplots
```{r}

```

### Making Sense of Data
```{r}
ggplot(data = Mitchell, aes(x = Month, y = Temp)) +
  geom_point() +
  scale_x_discrete(breaks = seq(0, 203, 12))
```

### A New Perspective
```{r}
# Stretch horizontal axis to reveal cyclic pattern
```

### Understanding Noise Age to Age Months
```{r}
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age) +
  geom_line()

head(pf.fc_by_age, 10)
pf.fc_by_age[17:19, ]

pf$age_with_months <- pf$age + (12 - pf$dob_month) / 12
```

### Age with Months Means
```{r}
library(dplyr)
pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age_with_months)

head(pf.fc_by_age_months)
```

### Noise in Conditional Means
```{r}
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age) +
  geom_line()

# Revise using age_with_months, subset by age_with_months < 71
ggplot(aes(x = age_with_months, y = friend_count_mean), 
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line()
```

### Smoothing Conditional Means
```{r}
# Save each plot to a variable, add smoothing lines, plot in one column
# Reducing number of bins reduces noise but may also leave out important details
# bias / variance trade-off

p1 <- ggplot(aes(x = age, y = friend_count_mean), 
             data = subset(pf.fc_by_age, age < 71)) +
  geom_line() +
  geom_smooth()

p2 <- ggplot(aes(x = age_with_months, y = friend_count_mean), 
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line() +
  geom_smooth()

p3 <- ggplot(aes(x = round(age / 5) * 5, y = friend_count), 
       data = subset(pf, age < 71)) +
  geom_line(stat = 'summary', fun.y = mean)

library(gridExtra)
grid.arrange(p2, p1, p3, ncol = 1)
```

