---
title: "EDA Lesson 5"
author: "Tom Gocken"
date: "Friday, January 29, 2016"
output: word_document
---

### Third Qualitative Variable
```{r}
# https://s3.amazonaws.com/udacity-hosted-downloads/ud651/pseudo_facebook.tsv
# Set directory, load data and ggplot2 package
setwd("C:/Projects/UD651")
list.files()
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
library(ggplot2)

# Boxplot with mean
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape = 4)

# Friend count by age, gender
ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(gender))) +
  geom_line(aes(color = gender), stat = 'summary', fun.y = median)

# Alt method by creating dataset in dplyr, chaining functions with %>%
library(dplyr)
pf.fc_by_age_gender <- pf %>%
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>%
  summarise(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n = n()) %>%
  ungroup() %>% # summarise ungroups gender layer, need to further ungroup age layer
  arrange(age)
```

### Plotting Conditional Summaries
```{r}
ggplot(aes(x = age, y = median_friend_count),
       data = pf.fc_by_age_gender) +
  geom_line(aes(color = gender))
```

### Thinking in Ratios
```{r}
# Creating dataset allows inspection of variables or further operations
```

### Reshaping Data
```{r}
# install.packages('reshape2')
library(reshape2)

pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = 'median_friend_count')

head(pf.fc_by_age_gender.wide)
```

### Ratio Plot
```{r}
# Plot females to males
ggplot(aes(x = age, y = female / male),
       data = pf.fc_by_age_gender.wide) +
  geom_line() +
  geom_hline(yintercept = 1, alpha = 0.3, linetype = 2)
```

### Third Quantitative Variable
```{r}
pf$year_joined <- floor(2014 - pf$tenure / 365)
```

### Cut a Variable
```{r}
summary(pf$year_joined)
table(pf$year_joined)

pf$year_joined.bucket <- cut(pf$year_joined,
                             c(2004, 2009, 2011, 2012, 2014))
```

### Plotting It All Together
```{r}
table(pf$year_joined.bucket, useNA = 'ifany')

ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(year_joined.bucket))) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary', fun.y = median)
```

### Plot the Grand Mean
```{r}
ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(year_joined.bucket))) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = mean, linetype = 2)
```

### Friending Rate
```{r}
with(subset(pf, tenure >= 1), summary(friend_count / tenure))
```

### Friendships Initiated
```{r}
ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)
```

### Bias Variance Trade off Revisited
```{r}
library(gridExtra)

p1 <- ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary', fun.y = mean)

p2 <- ggplot(aes(x = 7 * round(tenure/7), y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary', fun.y = mean)

p3 <- ggplot(aes(x = 30 * round(tenure/7), y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary', fun.y = mean)

p4 <- ggplot(aes(x = 90 * round(tenure/7), y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary', fun.y = mean)

# Using geom_smooth
ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_smooth(aes(color = year_joined.bucket))

grid.arrange(p1, p2, p3, p4, ncol = 1)
```

### Introducing the Yogurt Dataset
```{r}
# https://s3.amazonaws.com/udacity-hosted-downloads/ud651/yogurt.csv
```

### Histograms Revisited
```{r}
setwd("C:/Projects/UD651")
yo <- read.csv('yogurt.csv')
str(yo)

# Change id from int ot factor
yo$id <- factor(yo$id)
str(yo)

qplot(data = yo, x = price, fill = I('#F79420'))
```

### Number of Purchases
```{r}
summary(yo)
length(unique(yo$price))
table(yo$price)

str(yo)

yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)
```

### Prices Over Time
```{r}
qplot(x = all.purchases, data = yo, binwidth = 1, fill = I('#099DD9'))

ggplot(aes(x = time, y = price), data = yo) +
  geom_jitter(alpha = 1/4, shape = 21, fill = I('#F79420'))
```

### Looking at Samples of Households
```{r}
# Set seed for reproducible results
set.seed(4230)
# Sample 16 households
sample.ids <- sample(levels(yo$id), 16)

# Plot sample household purchases, facet by household
# %in% loops over ids
ggplot(aes(x = time, y = price),
       data = subset(yo, id %in% sample.ids)) +
  facet_wrap( ~ id) + 
  geom_line() + 
  geom_point(aes(size = all.purchases), pch = 1)
```

### Scatterplot Matrices
```{r}
# install.packages(GGally)
library(GGally)
theme_set(theme_minimal(20))

# Set seed for reproducible results
set.seed(1836)
pf_subset <- pf[, c(2:15)] # variables of interest
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000), ])
```

### Even More Variables
```{r}
# Micro-array/Gene Expression Data
# https://s3.amazonaws.com/udacity-hosted-downloads/ud651/nci.tsv
setwd("C:/Projects/UD651")
nci <- read.table('nci.tsv')
# Rename columns 1 to 64 for simpler labeling
colnames(nci) <- c(1:64)
```

### Heat Maps
```{r}
# Melt data to long format
library(reshape2)
nci.long.samp <- melt(as.matrix(nci[1:200, ]))
names(nci.long.samp) <- c('gene', 'case', 'value')
head(nci.long.samp)

# Plot first 200 genes
ggplot(aes(y = gene, x = case, fill = value),
       data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(c('blue', 'red'))(100))
```
